Global Distribution of Unintended Pregnancies and Contraceptive Usage

### 2024-03-15 code/eda-code/eda.qmd
Final Part 3 Updates
I organized the eda file and added some final interpretations. I cleaned the manuscript file to only include relevant information in the correct order and made sure the document renders.

```{r}
#load necessary packages
library(here) #for data loading/saving
library(dplyr) # for basic piping functions
library(skimr) # for summarizing data
library(ggplot2) # for visualizing data
library(gtsummary) # for creating tables
```

@@

@@

We also created a simple table with the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}
# isolate categorical variables into their own data set
mydata_cat <- mydata %>% select(region, larcster_usage, sarc_usage, trad_usage, none_usage)

@@

@@

We will make histogram plots for the continuous outcomes in our dataset.

It is important to consider the type of birth control methods associated with percent of unintended pregnancies as well. We will create a histogram for the `pct_upreg_sarc` variable.

@@

>>>>

```{r}
#| message: false
#load necessary packages
library(here) #for data loading/saving
library(dplyr) # for basic piping functions
library(skimr) # for summarizing data
library(ggplot2) # for visualizing data
library(gtsummary) # for creating tables
library(kableExtra) # for creating tables
```

@@

# Summarize the data 
```{r}
table(mydata$region)
table(mydata$subregion)
summary(mydata$pct_upreg)
ggplot(mydata, aes(x = country, y = pct_upreg, fill = region)) +
  geom_col(position = "dodge") +
  labs(title = "Percentage Unintended Pregnancies by Country",
       x = "Country",
       y = "Percentage")
```

The dataset includes 53 African countries, 36 Asian countries, 11 European Countries, 24 Latin American/Caribbean countries, and 8 Oceanic countries.There are 16 subregions that provide better detail about each country's location. The average percentage of unintended pregnancies at the country level is 46.79%. with a maximum of 75.65%. This variable has 24 missing values, which will need to be cleaned.  A simple visualization shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies.

@@

We created a simple table with the categorical variables: `region`, `larcster_usage`, `sarc_usage`, and `none_usage`.
```{r}
# isolate categorical variables into their own data set
mydata_cat <- mydata %>% select(region, larcster_usage, sarc_usage, trad_usage, none_usage)

@@

We also wanted a summary table with the predictor information to include in our manuscript.
```{r}
# Selecting variables
selected_variables <- mydata[c("pct_upreg_sarc", "region", "wra", "pct_currentlymarried", "pct_nevermarried","rate_matdeaths", "rate_matdeaths_upreg", "rate_matdeaths_abs")]

# Converting region from factor to numeric
selected_variables$region <- as.numeric(as.factor(selected_variables$region))

# Asia = 1
# Africa = 2
# Europe = 3
# Latin America & the Caribbean = 4
# Oceania = 5

# Create table 
kable(selected_variables, caption = "Table Depicting Predictor Variables", booktabs = TRUE) %>%
  kable_styling(full_width = F, font_size = 10)

# Save to file 
summarytable_file <- here("results", "tables", "predictortable.rds")
saveRDS(predictor_table, file = summarytable_file)
```

@@

We will make histogram plots for the continuous outcomes in our dataset. After exploring thepercent of unintended pregnancies related to each type of contraception method, we chose to focus on short acting reversible methods as our single outcome. All of the exploratory analysis on the other 3 potential outcomes can be found in the Supplementary-Material.qmd in the products > manuscript > supplement folder. The following exploratory analysis and subsequent statistical analysis will include the `pct_upreg_sarc` outcome only.

To begin exploring our outcome, we will create a histogram for the `pct_upreg_sarc` variable.

@@

We also need to explore the remaining predictor variables: `region`, `pct_currentlymarried`, `pct_nevermarried`, `rate_matdeaths`, `rate_matdeaths_upreg`, and `rate_matdeaths_abs`. Simple histograms will allow us to see the distribution of each variable and ensure there are no outliers. 
```{r}
# explore the region predictor
p22 <- mydata %>% ggplot(aes(x=region)) + 
  geom_bar() +
  ggtitle("Countries in Each Region") + 
  xlab("Region") + ylab("Number of Countries")
plot(p22)
figure_file = here("results", "figures", "exploratory-figures", "region_distribution.png")
ggsave(filename = figure_file, plot=p22)

# explore the current marriage predictor
p23 <- mydata %>% ggplot(aes(x=pct_currentlymarried)) + 
  geom_histogram() +
  ggtitle("Percent of Women Currently Married") + 
  xlab("Percentage") + ylab("Number of Countries")
plot(p23)
figure_file = here("results", "figures", "exploratory-figures", "currmarried_distribution.png")
ggsave(filename = figure_file, plot=p23)

# explore the never married predictor
p24 <- mydata %>% ggplot(aes(x=pct_nevermarried)) + 
  geom_histogram() +
  ggtitle("Percent of Women Never Married") + 
  xlab("Percentage") + ylab("Number of Countries")
plot(p24)
figure_file = here("results", "figures", "exploratory-figures", "nevmarried_distribution.png")
ggsave(filename = figure_file, plot=p24)

# explore the total maternal mortality rate predictor
p25 <- mydata %>% ggplot(aes(x=rate_matdeaths)) + 
  geom_histogram() +
  ggtitle("Total Rate of Maternal Deaths") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p25)
figure_file = here("results", "figures", "exploratory-figures", "matdeaths_distribution.png")
ggsave(filename = figure_file, plot=p25)

# explore outliers in maternal mortality from unintended pregnancies
mydata %>% filter(rate_matdeaths > 1000)

# explore the maternal mortality rate from unintended pregnancies predictor
p26 <- mydata %>% ggplot(aes(x=rate_matdeaths_upreg)) + 
  geom_histogram() +
  ggtitle("Rate of Maternal Deaths from Unintended Pregnancies") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p26)
figure_file = here("results", "figures", "exploratory-figures", "matdeathsupreg_distribution.png")
ggsave(filename = figure_file, plot=p26)

# explore outliers in maternal mortality from unintended pregnancies
mydata %>% filter(rate_matdeaths_upreg > 6000)

# explore the maternal mortality rate from abortions
p27 <- mydata %>% ggplot(aes(x=rate_matdeaths_abs)) + 
  geom_histogram() +
  ggtitle("Rate of Maternal Deaths from Abortions") + 
  xlab("Maternal Deaths") + ylab("Number of Countries")
plot(p27)
figure_file = here("results", "figures", "exploratory-figures", "matdeathsabs_distribution.png")
ggsave(filename = figure_file, plot=p27)

# explore outliers in maternal mortality from abortions
mydata %>% filter(rate_matdeaths_abs > 500)

```
P22: The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes. about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations.

P23: The distribution of women currently married is approximately normal with an average around 57% and no major outliers.

P24: The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%.

P25: The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality.

P26: The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category.

P27: The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor.

### 2024-03-15 code/processing-code/processing-file.qmd
Final Part 3 Updates
I organized the eda file and added some final interpretations. I cleaned the manuscript file to only include relevant information in the correct order and made sure the document renders.

```{r}
#| message: false
# path to data using here() function
data_location <- here("data","raw-data","AIU All Women Dataset.csv")
rawdata <- read_csv(data_location)
```

@@

We will load and examine the codebook first.

@@

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```
# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and contraception methods. We will use three predictor variables: total number of women using the contraceptive method, region, number of safe abortions,  . The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable with additional options for outcome variables of total percentage of unintended pregnancies by each contraceptive method: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). We will use multiple predictor variables including: total number of women using the contraceptive method, region, subregion, percentage of married women, percentage of women never married, total maternal deaths per 100,000 live births, deaths from unintended pregnancy per 100,000 live births, and deaths related to abortion per 100000 abortions. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be summed to create total number of women using each type of contraception: 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

>>>>

```{r}
#| message: false
# path to data using here() function
data_location <- here("data","raw-data","AIU All Women Dataset.csv")
rawdata <- read_csv(data_location)
```
There are 132 observations, one row for each low- to middle- income country, of the 399 variable included in the Adding It Up All Women Dataset. Some examples of numeric variables included are percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are three important character variables that can be used to group the numeric variables: country name, region, and sub region.

@@

We will load and examine the codebook to learn more about the dataset.

@@

```{r}
#| message: false
dplyr::glimpse(rawdata)
summary(rawdata)
head(rawdata)
skimr::skim(rawdata)
```

# Identifying variables to be cleaned 
We want to explore the relationship between unintended pregnancies and contraception methods. We will use three predictor variables: total number of women using the contraceptive method, region, number of safe abortions,  . The variable for the total percentage of unintended pregnancies, `pct_upreg` will be the main outcome variable with additional options for outcome variables of total percentage of unintended pregnancies by each contraceptive method: long acting reversible methods (`pct_upreg_larcster`), short term reversible methods (`pct_upreg_sarc`), traditional methods (`pct_upreg_trad`), and no method (`pct_upreg_nouse`). We will use multiple predictor variables including: total number of women using the contraceptive method, region, subregion, percentage of married women, percentage of women never married, total maternal deaths per 100,000 live births, deaths from unintended pregnancy per 100,000 live births, and deaths related to abortion per 100000 abortions. Covariates will include the number of women using each of these methods across four different age groups: 15-19, 20-24, 25-34, and 35-49. These variables will be summed to create total number of women using each type of contraception: 
-Number of women 15-19 using long-acting reversible methods `larc_1519`
-Number of women 15-19 using short-acting methods `sarc_1519`
-Number of women 15-19 using traditional methods `tradmethods_1519`
-Number of women 15-19 using no method `nomethod_1519`
-Number of women 20-24 using long-acting reversible methods `larc_2024`
-Number of women 20-24 using short-acting methods	`sarc_2024`
-Number of women 20-24 using traditional methods	`tradmethods_2024`
-Number of women 20-24 using no method	`nomethod_2024`
-Number of women 25-34 using long-acting reversible methods	`larc_2534`
-Number of women 25-34 using short-acting methods	`sarc_2534`
-Number of women 25-34 using traditional methods	`tradmethods_2534`
-Number of women 25-34 using no method	`nomethod_2534`
-Number of women 35-49 using long-acting reversible method	`larc_3549`
-Number of women 35-49 using short-acting method	`sarc_3549`
-Number of women 35-49 using all traditional methods	`tradmethods_3549`
-Number of women 35-49 using no method	`nomethod_3549`

### 2024-03-15 products/manuscript/Manuscript.qmd
Updated numeric scale on Predictor Variable table
Final Part 3 Updates
I organized the eda file and added some final interpretations. I cleaned the manuscript file to only include relevant information in the correct order and made sure the document renders.

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(readr)
library(knitr)
library(dplyr)
library(naniar)
library(ggplot2)
```
## Project Part 1

# Data Source 
The data source for this project is the Guttmacher Institute, which is a well-known research organization focusing on improving sexual and reproductive health around the world. The data used here is from their Adding It Up project. The goals of this project include estimating the need for, impact of, and costs associated with providing sexual and reproductive health services. Low and middle income countries are the target audience for this research, specifically women of reproductive age (15-49). Some of the variables included in this dataset are modern contraceptive use, unintended pregnancies, unplanned births, and abortions averted. Observations are drawn from nationally representative surveys, including Demographic and Health Surveys, UNICCEF Multiple Indicator Cluster Surveys, US Centers for Disease Control Reproductive Health Surveys, Performance Monitoring for Action Surveys, and others. The UN Population Division is the source of population projections for women in this age group, and estimates of unintended pregnancies are from the Guttmacher Institute, WHO, and other authors, which have been adjusted to 2019.

# Load and describe the data 
```{r}
library(here)
AIUdata <- read_csv(here("data", "raw-data", "AIU All Women Dataset.csv"))
dim(AIUdata)
str(AIUdata)
```
There are 132 observations, one row for each low- to middle- income country, of the 399 variable included in the Adding It Up All Women Dataset. Some examples of numeric variables included are percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are three important character variables that can be used to group the numeric variables: country name, region, and sub region.

# Summarize the data 
```{r}
table(AIUdata$region)
table(AIUdata$subregion)
summary(AIUdata$pct_upreg)
ggplot(AIUdata, aes(x = country, y = pct_upreg, fill = region)) +
  geom_col(position = "dodge") +
  labs(title = "Percentage Unintended Pregnancies by Country",
       x = "Country",
       y = "Percentage")
```
The dataset includes 53 African countries, 36 Asian countries, 11 European Countries, 24 Latin American/Caribbean countries, and 8 Oceanic countries.There are 16 subregions that provide better detail about each country's location. The average percentage of unintended pregnancies at the country level is 46.79%. with a maximum of 75.65%. This variable has 24 missing values, which will need to be cleaned.  A simple visualization shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies.

# Research Questions
1) Is there a significant difference in the percentage of unintended pregnancies among short acting reversible contraception users among countries?

# How to analyze 
Most of the predictors used in this analysis are numeric, so we can build linear regression models to analyze how each predictor affects the percentage of unintended pregnancies among women using short acting reversible methods. We can group countries together based on the subregion variable to compare how these relationships vary across the world as shown in the crude visualization above. There are other interesting variables that can be used for grouping for analysis such as wealth, urban/rural location, and marriage status.

@@

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across all age groups (15-19, 20-24, 25-35, and 35-49) to investigate the effectiveness of each contraceptive method. Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group. Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study. The table provided above under `Summarize the data` shows that African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists, but regulations differ. Chile, Colombia, and Ecuador legally recognize access, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision to victims of sexual and domestic violence (Hevia, 2012).

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._

This data originates from the Guttmacher Institute's Adding It Up project. The data contains numeric variables: percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are only three character variables that can be used to group the numeric variables: country name, region, and sub region. So far we have dvided the observations into "low", "medium", and "high" category of birth control usage.

@@

The next step is exploratory data analysis. The eda.qmd file in the code > eda-code folder contains all the code for this process. We produced a couple summary tables before we began a simple exploration of trends in percentage of unintended pregnancies by country in our first set of histograms. The second set of histograms explore contraceptive usage methods by country. These graphs revealed outliers, so we standardized by population size for each country. The next sets of plots are line graphs showing total usage of contraceptive methods to percentage of unintended pregnancies associated with that method. The last two sets of plots contain line graphs of the relationship between the two variables stratified by region and violin plots displaying the relationship between percentage of unintended pregnancies and the categorical usage variable created during processing. All of these figures are saved in the Supplementary-Material.qmd file in the products > manuscript > supplement folder. This process led us to focus on the percentage of unintended pregnancies among short acting reversible methods as our single outcome because this method is becoming increasingly popular, so we feel that results from this contraception method will be the most relevant.

@@

## Exploratory/Descriptive analysis (Project Part 2)
The first priority when cleaning the data was analyzing missing values, and our findings can be seen @tbl-summarytable. We found that the `pct_upreg` variable containing the percentage of unintended pregnancies is unreliable due to missing data that could not be explained. We decided to use the variables with percentage of unintended pregnancies related to each birth control method as our outcome variables because they are complete and were found directly based on the AIU protocol information. We printed the summary table here because it was useful in determining missingness and creating a new factor variable for the usage of each birth control method at the factor level later in the processing sequence. 
```{r}
#| label: tbl-summarytable
#| tbl-cap: "Summary table of numeric variables"
#| echo: FALSE
resulttable<- readRDS("../../results/tables/summarytable.rds")
kable(resulttable)
```

We explored the distribution of the countries by five categorical variables: `region`, `larcster_usage`, `sarc_usage`, `trad_usage`, and `none_usage` which is displayed in @tbl-summarytable2. We found that most countries are located in Africa followed by Asia and Latin America/the Caribbean. Only 20 countries have high usage of long acting reversible and sterilization methods compared to 33 countries with high usage of short acting reversible and traditional methods. It is interesting that the distributions for short acting reversible method, traditional method, and no method usages are the same. 
```{r}
#| label: tbl-summarytable2
#| tbl-cap: "Summary table for categorical variables"
#| echo: FALSE
resulttable2 <- readRDS("../../results/tables/summarytablecat.rds")
kable(resulttable2)
```

We created a multitude of exploratory plots to understand our data better. The histograms of percent of unintended pregnancies related to each method of birth control showed discrepancies in skewness which can be seen @exploratory-result. The `pct_upreg_larcster` and `pct_upreg_trad` variables are highly skewed right, which shows that more countries experience lower rates of unintended pregnancies when using these methods. The `pct_upreg_sarc` histogram is not as skewed. It shows that the percentage of unintended pregnancies among those using short acting reversible methods varies more between countries compared to long acting reversible methods. The histogram for `pct_upreg_none` shows that more countries report a higher percentage of unintended pregnancies among women who do not use any contraception, but it varies widely by country. 
```{r}
#| label: exploratory-result
#| fig-cap: "Histograms of Percentage of Unintended Pregnancy Use by Birth Control Method"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_larcster_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_trad_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_nouse_distribution.png"))
```

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies and the total number of women using that method of birth control for all three methods and those not using any method, which is to be expected. There is more variation among those using SARMs and traditional methods compared to those using long-acting reversible methods. Additionally, the lowest percentage in this category is around 35%, which is much higher than the percentages associated with the three birth control methods. These figures can be seen @exploratory-result2.
```{r}
#| label: exploratory-result2
#| fig-cap: "Relationship between Percentage of Unintended Pregnancies and Standardized Population Totals"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctuppreg_totallarcster.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totaltrad.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalnone.png"))
```

We stratified these four graphs by region to examine how the relationships differ across each of the 6 regions which is shown in @exploratory-result3 . The long acting reversible and sterilization methods graph is a little difficult to decipher, but Asia has the highest percentages of unintended pregnancies, followed by Latin America/the Caribbean and Africa. The graph for short acting reversible methods shows clearer differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa shows the sharpest increase in percentage of unintended pregnancies as the total population increases.Among women using traditional birth control methods, Europe has the sharpest increase in unintended pregnancies as population increases, but Asia has the highest percentage of unintended pregnancies in this category. The graph for women not using any birth control methods shows that Oceania has the highest percentage of unintended pregnancies. 
```{r}
#| label: exploratory-result3
#| fig-cap: "Percentage of Unintended Pregnancies Stratified by Region"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_larcster_stratified.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_trad_stratified.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_none_stratified.png"))
```

The last set of plots we created to explore the data are violin plots using the categorical usage variables for each birth control method as shown in @exploratory-result4. The first violin plot shows that the distribution of percentage of unintended pregnancies varies a lot among usage levels of long acting reversible and sterilization methods. High usage countries have a consistent percentage of unintended pregnancies, but countries with a medium usage level tend to have lower percentages of unintended pregnancies. It appears that the distribution of short acting reversible method usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method. Among women using traditional birth control methods, the distribution of percentage of unintended pregnancies is similar among medium and high usage countries, but low usage countries are skewed towards lower percentage of unintended pregnancies. While the distribution of percentage of unintended pregnancies is similar across the three usage levels for women not using any birth control method, the high usage countries appear to have a bottleneck in percentage of unintended pregnancies. The high and low usage countries appear to be skewed towards higher levels of unintended pregnancies, but the medium usage countries appear to have lower percentage of unintended pregnancies. 
```{r}
#| label: exploratory-result4
#| fig-cap: "Violin Plots with Birth Control Usage Levels"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "larcster_pctupreg_violin.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "trad_pctupreg_violin.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "none_pctupreg_violin.png"))
```

@@

We completed a simple linear regression analysis with each of the 7 predictors. The predictor with the strongest relationship with the outcome is total number of women using short acting reversible methods (`sarc_standard`).

```{r}
#| label: basic-statistical-analysis
#| fig-cap: "Strongest Predictor: sarc_standard"
#| echo: FALSE
readRDS(here("results","tables", "resulttable1.rds")) 
```
The model using the region predictor revealed that the Latin America/the Caribbean region had the strongest relationship with the outcome variable followed closely by the Asia region. The Oceania region does not appear to have a strong effect on the outcome variable with the Europe region only showing a slightly larger association.
```{r}
#| label: basic-statistical-analysis
#| fig-cap: "Regional Predictor"
#| echo: FALSE
readRDS(here("results","tables", "resulttable2.rds")) 
```
We created one model with all 7 predictors using multiple linear regression. This model shows the strongest predictor is the number of women using short acting reversible contraception. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. Both of the marriage status predictors are associated with a decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both flipped from the single predictor models to show a positive association with the outcome. 
```{r}
#| label: basic-statistical-analysis
#| fig-cap: "Multiple Linear Regression with 7 Predictors"
#| echo: FALSE
readRDS(here("results","tables", "resulttable8.rds")) 
```
The full statistical analysis can be found in the statistical-analysis.qmd file in the code > analysis-code folder.

>>>>

```{r, echo=FALSE, message=FALSE}
# load a few R packages
library(here)
library(readr)
library(knitr)
library(dplyr)
library(naniar)
library(ggplot2)
```

@@

We will be looking for patterns in contraceptive methods and higher unintended pregnancies across all age groups (15-19, 20-24, 25-35, and 35-49) to investigate the effectiveness of three contraceptive methods: long acting reversible and sterilization,  short acting reversible, and traditional. Financial, social, religious, and cultural factors are current challenges against efforts towards reducing unintended pregnancies. According to a cross-sectional survey done in 2017, contraceptive use was prevalent among women who had attained higher education, were in stable partnerships, and identified with the dominant ethnic group. Additionally, factors like previous pregnancies and immigration status did not appear to influence contraceptive use in this study. African and Latin American/Caribbean countries have the highest percentages of unintended pregnancies. This could be due to the current legal status of contraception within Latin American countries. Free emergency contraception access exists, but regulations differ. Chile, Colombia, and Ecuador legally recognize access, while Nicaragua and Bolivia rely on Ministerial protocols. Argentina and Brazil lack legal recognition, but offer protocols and guides. Mexico requires provision to victims of sexual and domestic violence (Hevia, 2012).

## Description of data and data source
_Describe what the data is, what it contains, where it is from, etc. Eventually this might be part of a methods section._
The data source for this project is the Guttmacher Institute, which is a well-known research organization focusing on improving sexual and reproductive health around the world. The data used here is from their Adding It Up project. The goals of this project include estimating the need for, impact of, and costs associated with providing sexual and reproductive health services. Low and middle income countries are the target audience for this research, specifically women of reproductive age (15-49). Some of the variables included in this dataset are modern contraceptive use, unintended pregnancies, unplanned births, and abortions averted. Observations are drawn from nationally representative surveys, including Demographic and Health Surveys, UNICCEF Multiple Indicator Cluster Surveys, US Centers for Disease Control Reproductive Health Surveys, Performance Monitoring for Action Surveys, and others. The UN Population Division is the source of population projections for women in this age group, and estimates of unintended pregnancies are from the Guttmacher Institute, WHO, and other authors, which have been adjusted to 2019.

For the purpose of this project, we chose several numeric variables from the data source: percentages of pregnancy outcomes, rates of safe and unsafe abortions, percentage of various contraceptive usages, percentages of care received for various pregnancy complications, and costs for many variables including abortions and STIs. There are only three character variables that can be used to group the numeric variables: country name, region, and sub region.

@@

The next step is exploratory data analysis. The eda.qmd file in the code > eda-code folder contains all the code for this process. We produced a couple summary tables before we began a simple exploration of trends in percentage of unintended pregnancies by country in our first set of histograms. The second set of histograms explore contraceptive usage methods by country. These graphs revealed outliers, so we standardized by population size for each country. The next sets of plots are line graphs showing total usage of contraceptive methods to percentage of unintended pregnancies associated with that method. The last two sets of plots contain line graphs of the relationship between the two variables stratified by region and violin plots displaying the relationship between percentage of unintended pregnancies and the categorical usage variable created during processing. All of these figures are saved in the Supplementary-Material.qmd file in the products > manuscript > supplement folder. This process led us to focus on the percentage of unintended pregnancies among short acting reversible methods as our single outcome because this method is becoming increasingly popular, so we feel that results from this contraception method will be the most relevant. Additionally, the exploratory data analysis process showed that the distribution of this outcome variable is less skewed than the other options, and the stratification by region is clearer, which allows for better understanding.

@@

## Exploratory/Descriptive analysis

After deciding on a single outcome variable, we created a summary table with that outcome and the seven predictors we will use to predict it, which can be seen @tbl-summarytable.
```{r}
#| label: tbl-summarytable
#| tbl-cap: "Summary table of numeric variables"
#| echo: FALSE
predictortable<- readRDS(here("results", "tables", "predictortable.rds"))
predictortable
```

We created a multitude of exploratory plots to understand our data better. The first @exploratory-result shows the distribution of the outcome variable: percent of unintended pregnancies due to short acting reversible methods. The `pct_upreg_sarc` variable is slightly skewed to the right with an average around 12%.
```{r}
#| label: exploratory-result
#| fig-cap: "Histograms of Percentage of Unintended Pregnancy Use by Birth Control Method"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_distribution.png"))
```

The next set of variables we explored were the total counts of women using each birth control method. There were two large outliers that prevented the graphs from being useful. After exploring the outliers, we standardized using the population size of reproductive aged women (15-49 years). There is a positive association between the percentage of unintended pregnancies among short acting reversible method users and the total number of women using that method of birth control, which is to be expected. This association is displayed in @exploratory-result2.
```{r}
#| label: exploratory-result2
#| fig-cap: "Relationship between Percentage of Unintended Pregnancies and Standardized Population Total"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_totalsarc.png"))
```

We stratified these four graphs by region to examine how the relationship differs across each of the t regions which is shown in @exploratory-result3 . The graph for short acting reversible methods shows clearer differences between regions, and Latin America/the Caribbean leads in the percentage of unintended pregnancies in this category, but Africa shows the sharpest increase in percentage of unintended pregnancies as the total population increases.
```{r}
#| label: exploratory-result3
#| fig-cap: "Percentage of Unintended Pregnancies Stratified by Region"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "pctupreg_sarc_stratified.png"))
```

We also created violin plots using the categorical usage variables for each birth control method as shown in @exploratory-result4. It appears that the distribution of short acting reversible method usage is skewed towards lower percentages of unintended pregnancies across all usage levels, suggesting that more countries have lower percentage of unintended pregnancies among women using this method. The high usage countries appear to have a bottleneck in percentage of unintended pregnancies. The high and low usage countries appear to be skewed towards higher levels of unintended pregnancies, but the medium usage countries appear to have lower percentage of unintended pregnancies. 
```{r}
#| label: exploratory-result4
#| fig-cap: "Violin Plots with Birth Control Usage Levels"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "sarc_pctupreg_violin.png"))
```

Lastly, we explored the predictor variables that will be used in our models, and the graphs are shown in @exploratory-result5. The Africa region accounts for most of the observations with about 53 countries, followed by Asia that includes. about 35 countries. Latin America/the Caribbean accounts for about 25 countries. Europe and Oceania account for the least number of observations. The distribution of women currently married is approximately normal with an average around 57% and no major outliers. The distribution of women never married is slightly skewed to the right with an average around 30%, but there are a few outliers greater than 60%. The distribution of the total maternal mortality rate is highly skewed to the right with a couple outliers. We would have expected the highest rates to belong to India and China because they have the most people, but it is Chad, Sierra Leone, and South Sudan that account for the highest rates of total maternal mortality. The distribution of the maternal mortality rate from unintended pregnancies is highly skewed to the right with a couple outliers, which is to be expected based on the trend shown in the previous histogram. Chad and South Sudan also account for the highest rates of maternal mortality in this category. The distribution of maternal mortality rate from abortions is also highly skewed to the right. The outliers are accounted for by Gambia and Mauritania in this predictor.
```{r}
#| label: exploratory-result5
#| fig-cap: "Predictor Variable Distributions"
#| echo: FALSE
knitr::include_graphics(here("results", "figures", "exploratory-figures", "region_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "region_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "currmarried_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "nevmarried_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "matdeaths_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "matdeathsupreg_distribution.png"))
knitr::include_graphics(here("results", "figures", "exploratory-figures", "matdeathsabs_distribution.png"))
```

@@

We completed a simple linear regression analysis with each of the 7 predictors shown in @statistical-analysis1. The predictor with the strongest relationship with the outcome is total number of women using short acting reversible methods (`sarc_standard`).

```{r}
#| label: statistical-analysis1
#| fig-cap: "Strongest Predictor: sarc_standard"
#| echo: FALSE
readRDS(here("results","tables", "resulttable1.rds")) 
```

The second model we created used the region predictor, which revealed that the Latin America/the Caribbean region had the strongest relationship with the outcome variable followed closely by the Asia region. The Oceania region does not appear to have a strong effect on the outcome variable with the Europe region only showing a slightly larger association. These associations can be seen in @statistical-analysis2.
```{r}
#| label: statistical-analysis2
#| fig-cap: "Regional Predictor"
#| echo: FALSE
readRDS(here("results","tables", "resulttable2.rds")) 
```
We created one model with all 7 predictors using multiple linear regression. This model shows the strongest predictor is the number of women using short acting reversible contraception. The next strongest predictor is the European region followed by the Latin America/the Caribbean region and the Asia region, which is a shift from the model with the singular region predictor. The Oceania region now has a negative association with the outcome. Both of the marriage status predictors are associated with a decrease in the outcome. The coefficient estimates for the rate of maternal death due to unintended pregnancies and abortion both flipped from the single predictor models to show a positive association with the outcome. 
```{r}
#| label: basic-statistical-analysis
#| fig-cap: "Multiple Linear Regression with 7 Predictors"
#| echo: FALSE
readRDS(here("results","tables", "resulttable8.rds")) 
```
The full statistical analysis can be found in the statistical-analysis.qmd file in the code > analysis-code folder.
